TADCompare: an R package for differential analysis and characterization of Topologically Associated Domains

### 2019-09-19 manuscript/manuscript.Rmd
Search and replace trait to phenotype, mainly in Methods

**Conclusions.** We present a general-purpose method for estimating and comparing the coherence of phenotype-associated gene networks that accounts for the network size and shape differences. Our results highlight the gaps in of our current knowledge of the genetics and molecular mechanisms of complext traits and suggest priorities for future work.

@@

Trait-associated gene networks were formed by mapping trait-associated genes to PPIs. Genes without PPI annotations (e.g., C5orf4, LOC100507462), antisense transcripts (e.g., CDKN2B-AS), readthrough (e.g., NPHP3-ACAD11), non-protein-coding transcripts (e.g., LINC00478) were excluded. Genes that do not have PPIs with each other but interact with other genes in the global PPI network were kept. Traits with less than five genes were omitted. Additionally, traits with genes having identical internal degrees (hence, no degree distribution) were filtered due to the inability to derive a fit through the internal vs. external degree distributions (see below). These precautions were set to avoid situations where a small number of genes in a network can cause instability in the results. Importantly, the number of trait-associated genes having PPI annotations depends on the content of a PPI database; hence, some traits may be filtered when using one but not the other PPI database to build trait-associated networks.

@@

Trait-specific **internal degree distribution** was defined as the number of interactions between trait-associated genes, or, more precisely, their interacting protein products. As different traits have different numbers of associated genes forming networks of unknown configurations, the corresponding internal degree distribution was normalized to the expected interaction count for a network of that size (see Results).

Trait-specific **external degree distribution** was defined as the number of interactions between genes in a trait-associated network and all other genes in the PPI network. Being measured in the context of the entire PPI network of fixed size, the external degree distribution did not require normalization.

To compare trait-specific degree distributions, the normalized internal vs. external degree distributions were plotted on a scatterplot. The square root of the degree was taken for more informative representation. For each trait, the internal vs. external degree distribution plot was fit with a regression line through the origin to capture the slopes. Slopes were used as a measure of coherence with lower slopes being associated with higher coherence. Degree distributions of the network nodes were obtained using the `igraph` R package v. 1.0.1.

@@

The slopes of the internal vs. external degree distributions of the networks exhibiting high and random coherence (KEGG and random networks, respectively)  are dependent on the content of a PPI database. Thus, we normalized the slopes for the trait-associated networks using slopes formed by the degree distributions of similarly sized networks with high coherence ($\beta_{min}$ slope of KEGG networks) and random coherence ($\beta_{max}$ slope of random networks). For each trait $i$ with slope $\beta_i$, we calculate the normalized slope $\beta_{i\ norm} = \frac{\beta_{i} - \beta_{max}}{\beta_{min}- \beta_{max}}$, which gives a measure of coherence on the range of [0,1]. Higher values indicate a higher level of coherence similar to that of KEGG pathways. These values represent normalized coherence and can be compared across traits and PPI databases.

@@

To test whether each slope is significantly different than random, we used a permutation test. For trait $i$ with the gene network of size $g_i$, we define $\beta_i$ as the slope derived by regressing the external degree distribution on the internal degree distribution. We then generate $n=10000$ networks of randomly selected genes of the same size $g_i$ as the set for trait $i$. For each random network $j$, we identify its slope $\beta_{r}$ representing the internal vs. external degree distribution regression line of random coherence. We calculate the permutation p-value as $p_i = \frac{(\sum_{j=1}^nI({\beta_{rj} > \beta_i}) + 1}{n + 1}$ [@Phipson:2010aa]. Intuitively, this approach estimates whether the slope of the trait-specific network $\beta_i$ is consistently larger than those of random networks. All analyses and visualizations were performed in the R/Bioconductor computing environment v.3.4.0 [@gentleman2006bioinformatics].

>>>>

**Conclusions.** We present a general-purpose method for estimating and comparing the coherence of phenotype-associated gene networks that accounts for the network size and shape differences. Our results highlight the gaps in of our current knowledge of the genetics and molecular mechanisms of complext phenotypes and suggest priorities for future work.

@@

Phenotype-associated gene networks were formed by mapping trait-associated genes to PPIs. Genes without PPI annotations (e.g., C5orf4, LOC100507462), antisense transcripts (e.g., CDKN2B-AS), readthrough (e.g., NPHP3-ACAD11), non-protein-coding transcripts (e.g., LINC00478) were excluded. Genes that do not have PPIs with each other but interact with other genes in the global PPI network were kept. Phenotypes with less than five genes were omitted. Additionally, phenotypes with genes having identical internal degrees (hence, no degree distribution) were filtered due to the inability to derive a fit through the internal vs. external degree distributions (see below). These precautions were set to avoid situations where a small number of genes in a network can cause instability in the results. Importantly, the number of phenotype-associated genes having PPI annotations depends on the content of a PPI database; hence, some phenotypes may be filtered when using one but not the other PPI database to build phenotype-associated networks.

@@

Phenotype-specific **internal degree distribution** was defined as the number of interactions between phenotype-associated genes, or, more precisely, their interacting protein products. As different phenotypes have different numbers of associated genes forming networks of unknown configurations, the corresponding internal degree distribution was normalized to the expected interaction count for a network of that size (see Results).

Phenotype-specific **external degree distribution** was defined as the number of interactions between genes in a phenotype-associated network and all other genes in the PPI network. Being measured in the context of the entire PPI network of fixed size, the external degree distribution did not require normalization.

To compare phenotype-specific degree distributions, the normalized internal vs. external degree distributions were plotted on a scatterplot. The square root of the degree was taken for more informative representation. For each phenotype, the internal vs. external degree distribution plot was fit with a regression line through the origin to capture the slopes. Slopes were used as a measure of coherence with lower slopes being associated with higher coherence. Degree distributions of the network nodes were obtained using the `igraph` R package v. 1.0.1.

@@

The slopes of the internal vs. external degree distributions of the networks exhibiting high and random coherence (KEGG and random networks, respectively)  are dependent on the content of a PPI database. Thus, we normalized the slopes for the phenotype-associated networks using slopes formed by the degree distributions of similarly sized networks with high coherence ($\beta_{min}$ slope of KEGG networks) and random coherence ($\beta_{max}$ slope of random networks). For each phenotype $i$ with slope $\beta_i$, we calculate the normalized slope $\beta_{i\ norm} = \frac{\beta_{i} - \beta_{max}}{\beta_{min}- \beta_{max}}$, which gives a measure of coherence on the range of [0,1]. Higher values indicate a higher level of coherence similar to that of KEGG pathways. These values represent normalized coherence and can be compared across phenotypes and PPI databases.

@@

To test whether each slope is significantly different than random, we used a permutation test. For phenotype $i$ with the gene network of size $g_i$, we define $\beta_i$ as the slope derived by regressing the external degree distribution on the internal degree distribution. We then generate $n=10000$ networks of randomly selected genes of the same size $g_i$ as the set for phenotype $i$. For each random network $j$, we identify its slope $\beta_{r}$ representing the internal vs. external degree distribution regression line of random coherence. We calculate the permutation p-value as $p_i = \frac{(\sum_{j=1}^nI({\beta_{rj} > \beta_i}) + 1}{n + 1}$ [@Phipson:2010aa]. Intuitively, this approach estimates whether the slope of the phenotype-specific network $\beta_i$ is consistently larger than those of random networks. All analyses and visualizations were performed in the R/Bioconductor computing environment v.3.4.0 [@gentleman2006bioinformatics].
